\section{Introduction}

For many reasons, the need for communication has become one of the most urgent
problems in the modern world. Nowadays, smartphones are almost universally used
as the primary form of communication. As a result, it would be incredible if
they could communicate when there is no infrastructure since then, people
wouldn't always be dependent on it. Anytime they want, they can talk.
Therefore, it has also become more and more essential to communicate
efficiently in places devoid of infrastructure. There are many well-known
smartphone chat applications, including WhatsApp, Telegram, Viber, and others;
nevertheless, they all depend on the internet and the appropriate communication
infrastructure.

When we say that an infrastructure is available, we mean that base stations or
access points (such as cellular towers or Wi-Fi routers) exist that the nodes
can connect to and communicate with. But when it comes to an
infrastructureless environment, the nodes don’t have to connect to the
cellular towers since they establish direct wireless connections. In this case,
the ability to leverage various infrastructure-free communication protocols is
referred to as being adaptive. Thus, mobile ad-hoc networks are networks
created dynamically without the need for infrastructure, presuming that all
nodes connecting to the network will be portable to some
extent\cite{chlamtac2003}.

Wi-Fi (in ad-hoc mode) and Bluetooth come out on top regarding available
infrastructureless communication methods. Bluetooth is one of the most popular
short-range wireless technologies for transferring data between devices over
short distances. It took the place of the conventional cable connections that
were used to transport data between our regularly used portable gadgets. When
compared to Wi-Fi, Bluetooth is less expensive and uses less power, but Wi-Fi
offers more benefits, such as faster speeds and broader transmission ranges.
Although Wi-Fi's ad hoc mode could handle direct device-to-device (D2D)
communication without needing an Access Point (AP), it was not generally used
because of some issues. On the other hand, Wi-Fi Direct (Wi-Fi P2P), which is
constructed on top of the infrastructure mode of Wi-Fi, enables simple and
effective D2D communication and lets the devices choose who will take over the
AP-like functions. Another technology to consider is Bluetooth Low Energy
(BLE), a development of Bluetooth that significantly reduces battery usage
because it establishes brief connections and stays in sleep mode when not
connected.

When putting out a proposal for an infrastructure-less communication solution,
a few more crucial factors need to be taken into account. Affordable power
consumption, message transmission dependability, message delays, routing
mechanism performance in these scenarios, and network establishment speed in
scenarios without infrastructure are a few of them. The topology of mobile ad
hoc networks frequently changes due to factors including node mobility and
unexpected node unavailability.
Due to this, it is crucial to find appropriate routing methods and neighbor
identification techniques to increase communication effectiveness in these
kinds of networks.

Meshify is a framework for infrastructure-less communication using smartphones.
It is currently equipped only with Bluetooth classic and BLE
technologies\cite{gunasekara2022}. Implementing Wi-Fi Direct as an additional
wireless communication technology for Meshify will significantly enhance its
connectivity capabilities. Wi-Fi Direct offers faster data transfer speeds and
a more stable connection than Bluetooth, making it ideal for real-time data
exchange applications, as explained above. This expansion will empower Meshify
with a versatile and robust communication framework, improving its performance
and usability. Also, improvements should focus on improving the current Meshify
neighbor discovery protocol for accurate device detection and multi-hop message
relaying to extend network reach. Incorporating research from opportunistic
networks can enhance communication resilience. In addition, the Meshify
middleware framework should adhere to the best software engineering principles
and design patterns to ensure that third-party app developers may readily use
it.
